Decoding UK Hospital Wait Times: Infrastructure, Workforce, and Demand

### 2024-11-24 paper/paper.qmd
Added data citation and package citations

---
title: "Decoding UK Hospital Wait Times: Infrastructure, Workforce, and Demand"
subtitle: "How Systemic Pressures Shape Delays in Medical Care"
author: 
  - Chenika Bukes
thanks: "Code and data are available at: [https://github.com/chenikabukes/UnitedKingdomHealthcare](https://github.com/chenikabukes/UnitedKingdomHealthcare)."
date: today
date-format: long
abstract: "Hospital wait times in the United Kingdom are a critical measure of healthcare efficiency, reflecting the balance between resources and demand. This study examines the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for seven key medical procedures from 2015 to 2019. Using regression modeling, we find that reduced hospital beds per capita significantly prolong wait times,and while higher physician availability helps, it does little to offset the impact of declining bed availability. Additionally, Type 1 major care attendance rates correlate with increased delays, further indicating systemic bottlenecks. These findings highlight the urgent need for strategic resource allocation and demand management to address delays and improve patient outcomes."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Understanding the drivers of wait times is vital for improving healthcare access and efficiency. By identifying the factors contributing to delays, this paper offers actionable insights for policymakers and healthcare administrators seeking to enhance patient outcomes and system resilience. The remainder of this paper is structured as follows: Section 1 details the data and methodology, Section 2 presents key results, Section 3 discusses implications, and Section 4 concludes with recommendations for future research.

@@

In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR], using the haven and `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al. 2021),. All figures in the report are generated using `ggplot2` [@citeGG]. We run the model in R [@citeR] using the `modelsummary` package of [@modelsummary].

@@

The NHS A&E Attendances and Emergency Admissions dataset captures the monthly demand for emergency services. It includes the total number of attendances at Accident & Emergency (A&E) departments and emergency admissions, alongside measures of wait times for admission. These statistics, collected at the provider organization level, are drawn from NHS Trusts, Foundation Trusts, and independent sector organizations. The data was aggregated from monthly submissions, a change implemented after Sir Bruce Keogh’s review of waiting time standards. The dataset includes attendance and admission figures indexed to pre-pandemic and earlier baseline levels, enabling us to monitor trends and variations across years.

The remaining datasets are sourced from the OECD Health Statistics 2024 database, which provides standardized, internationally comparable data across healthcare systems. The OECD collaborates with organizations such as the United Nations and Eurostat to develop robust methodologies and benchmarks for health statistics. For the United Kingdom, data on hospital resources (e.g., hospital counts and bed availability), physician counts, and other health system metrics are provided by various entities like NHS Digital and Public Health Scotland. The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments.

@@

**Overall**: All datasets rely on administrative data validated by their respective agencies, minimizing reporting errors. While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not materially affect the study's conclusions. Similar datasets from organizations like WHO or Eurostat could offer supplementary insights but were not utilized due to a lack of UK-specific detail and consistency.

## Methodology #{sec-methodology}
EXPLAIN HOW YOU CLEANED DATA HERE

@@

@@

ggplot(cleaned_data, aes(x = Year, y = physicians_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Physicians per 1 000 rates from 2015 to 2019", x = "Year", y = "Physicians per 1 000") +
  theme_minimal()
```
@fig-physicians shows a steady increase in the number of physicians per 1,000 inhabitants from 2015 to 2019, rising from approximately 2.0 to 2.15. This consistent growth indicates an improvement in physician availability relative to population size, potentially reflecting increased investment in healthcare workforce development or recruitment efforts. The trend suggests progress in addressing healthcare access, but further analysis is needed to determine whether this growth aligns with demand and regional healthcare needs.

ggplot(cleaned_data, aes(x = Year, y = beds_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2019", x = "Year", y = "Beds per 1 000 Inhabitants") +
  theme_minimal()
```
@fig-beds shows a steady decline in hospital beds per 1,000 inhabitants from 2015 to 2019, decreasing from approximately 2.6 to 2.45 beds. This consistent downward trend suggests a reduction in bed capacity relative to population growth, potentially indicating healthcare system pressures such as resource reallocation, efficiency measures, or underinvestment in infrastructure. The decline highlights the need to assess the impact on patient care and explore strategies to balance population needs with available resources.

@@

# Plotting Type 1 Major Care Attendance
ggplot(cleaned_data) +
  geom_line(aes(x = Year, y = attendance, color = "A&E Attendances"), size = 1) +
  scale_color_manual(values = c("A&E Attendances" = "blue")) +
  labs(
    title = "Type 1 Major Care Attendance in the United Kingdom (2015–2019)",
    x = "Year",
    y = "Percentage Change from 2015 (%)",
    color = "Service Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```
@fig-attendance-demand illustrates the percentage change in Type 1 Major Care Attendance (A&E Attendances) in the UK from 2015 to 2019. A&E attendances showed a modest, steady increase over the observed period, plateauing slightly by 2019. This trend highlights the growing demand for emergency services and underscores the need for further investigation into factors contributing to increased attendance, such as population growth, access to primary care, or changes in healthcare-seeking behavior. Addressing these factors through improved preventive and community-based care could help reduce the burden on emergency services.

# Model
After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., number of physicians, hospital beds per 1,000 people, and attendance rates) and wait times for medical procedures. These variables show potential predictive power, indicating a linear relationship. To further investigate and make predictions about how these factors influence wait times, we will implement a linear regression model.

@@

\begin{align} 
$Y_{ij}$ is the average percenta change in wait time for the 7 specific medical procedures in year $i$.\\
$X_{1i}, X_{2i}, X_{3i}$ represent the independent variables of physicians per 1000 people, beds per 1000 people, Type 1 attendance for procedure $j$ in year $i$.\\
$\beta_0$ is the intercept or constant term, representing the expected wait time when all predictors are zero.\\
$\beta_1, \beta_2, \beta_3$ are the slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.\\
$\epsilon_{i}$ is the error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.
\end{align}

@@

# Results

@@

# Discussion

>>>>

---
title: "Decoding UK Hospital Wait Times: Infrastructure, Workforce, and Demand"
subtitle: "How Systemic Pressures Shape Delays in Medical Care"
author: 
  - Chenika Bukes
thanks: "Code and data are available at: [https://github.com/chenikabukes/UnitedKingdomHealthcare](https://github.com/chenikabukes/UnitedKingdomHealthcare)."
date: today
date-format: long
abstract: "Hospital wait times in the United Kingdom are a critical measure of healthcare efficiency, reflecting the balance between resources and demand. This study examines the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for seven key medical procedures from 2015 to 2019. Using regression modeling, we find that reduced hospital beds per capita significantly prolong wait times,and while higher physician availability helps, it does little to offset the impact of declining bed availability. Additionally, Type 1 major care attendance rates correlate with increased delays, further indicating systemic bottlenecks. These findings highlight the urgent need for strategic resource allocation and demand management to address delays and improve patient outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

Understanding the drivers of wait times is vital for improving healthcare access and efficiency. By identifying the factors contributing to delays, this paper offers actionable insights for policymakers and healthcare administrators seeking to enhance patient outcomes and system resilience. The remainder of this paper is structured as follows: @sec-data details the data and methodology, @sec-model presents the model used for analysis of the data, @sec-results presents the key results from the model, and @sec-discussion concludes with final analysis of the results and recommendations for future research.

@@

In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR], using the haven and `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al. 2021). All figures in the report are generated using `ggplot2` [@citeGG]. We run the model in R [@citeR] using the `modelsummary` package of [@modelsummary].

@@

The NHS A&E Attendances and Emergency Admissions dataset [@attendance] captures the monthly demand for emergency services. It includes the total number of attendances at Accident & Emergency (A&E) departments and emergency admissions, alongside measures of wait times for admission. These statistics, collected at the provider organization level, are drawn from NHS Trusts, Foundation Trusts, and independent sector organizations. The data was aggregated from monthly submissions, a change implemented after Sir Bruce Keogh’s review of waiting time standards. The dataset includes attendance and admission figures indexed to pre-pandemic and earlier baseline levels, enabling us to monitor trends and variations across years.

The remaining datasets are sourced from the OECD Health Statistics 2024 database [@health], which provides standardized, internationally comparable data across healthcare systems. The OECD collaborates with organizations such as the United Nations and Eurostat to develop robust methodologies and benchmarks for health statistics. For the United Kingdom, data on hospital resources (e.g., hospital counts and bed availability), physician counts, and other health system metrics are provided by various entities like NHS Digital and Public Health Scotland. The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments.

@@

**Overall**: All datasets rely on administrative data validated by their respective agencies, minimizing reporting errors. While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not affect the study's time period from 2015 to 2019. Similar datasets from organizations like WHO or Eurostat could offer supplementary insights but were not utilized due to a lack of UK-specific detail and consistency.

## Methodology {#sec-methodology}
EXPLAIN HOW YOU CLEANED DATA HERE

@@

ggplot(cleaned_data, aes(x = Year, y = beds_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2019", x = "Year", y = "Beds per 1 000 Inhabitants") +
  theme_minimal()
```

@fig-beds shows a steady decline in hospital beds per 1,000 inhabitants from 2015 to 2019, decreasing from approximately 2.6 to 2.45 beds. This consistent downward trend suggests a reduction in bed capacity relative to population growth, potentially indicating healthcare system pressures such as resource reallocation, efficiency measures, or underinvestment in infrastructure. The decline highlights the need to assess the impact on patient care and explore strategies to balance population needs with available resources.

@@

ggplot(cleaned_data, aes(x = Year, y = physicians_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Physicians per 1 000 rates from 2015 to 2019", x = "Year", y = "Physicians per 1 000") +
  theme_minimal()
```

@fig-physicians shows a steady increase in the number of physicians per 1,000 inhabitants from 2015 to 2019, rising from approximately 2.0 to 2.15. This consistent growth indicates an improvement in physician availability relative to population size, potentially reflecting increased investment in healthcare workforce development or recruitment efforts. The trend suggests progress in addressing healthcare access, but further analysis is needed to determine whether this growth aligns with demand and regional healthcare needs.

@@

# Plotting Type 1 Major Care Attendance
ggplot(cleaned_data) +
  geom_line(aes(x = Year, y = attendance, color = "A&E Attendances"), size = 1) +
  scale_color_manual(values = c("A&E Attendances" = "blue")) +
  labs(
    title = "Type 1 Major Care Attendance in the United Kingdom (2015–2019)",
    x = "Year",
    y = "Percentage Change from 2015 (%)",
    color = "Service Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
@fig-attendance-demand illustrates the percentage change in Type 1 Major Care Attendance (A&E Attendances) in the UK from 2015 to 2019. A&E attendances showed a modest, steady increase over the observed period, plateauing slightly by 2019. This trend highlights the growing demand for emergency services and underscores the need for further investigation into factors contributing to increased attendance, such as population growth, access to primary care, or changes in healthcare-seeking behavior. Addressing these factors through improved preventive and community-based care could help reduce the burden on emergency services.

# Model {#sec-model}
After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., number of physicians, hospital beds per 1,000 people, and attendance rates) and wait times for medical procedures. These variables show potential predictive power, indicating a linear relationship. To further investigate and make predictions about how these factors influence wait times, we will implement a linear regression model.

@@

\begin{align} 
Y_{ij} & = \text{The average percentage change in wait time for the 7 specific medical procedures in year } i. \\
X_{1i}, X_{2i}, X_{3i} & = \text{The independent variables of physicians per 1000 people, beds per 1000 people, and Type 1 attendance for procedure } j \text{ in year } i. \\
\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \\
\beta_1, \beta_2, \beta_3 & = \text{The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.} \\
\epsilon_{i} & = \text{The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.}
\end{align}

@@

# Results {#sec-results}

@@

# Discussion {#sec-discussion}
